Mergeable Types

### 2017-02-27 paper/examples.tex
Rope Edits.

@@

Fig.~\ref{fig:rope-sig} shows a rope interface in OCaml. The type
of the list of values is left abstract (\C{A.t}); it can be
instantiated with any module that supports \C{concat} and \C{split}
operations. The interface shown in Fig.~\ref{fig:rope-sig} is that of
a mergeable rope (\C{MRope});

@@

{\bf Lists}. As an abstract data type, lists support three
operations: \C{insert x i l}, to insert an element \C{x} at position
\C{i} in the list \C{l}, \C{delete i l}, to delete the element at
\C{i}'th position, and \C{subst i x l}, to substitute the element at
the \C{i}'th position with \C{x}.  Lists of mergeable items are
mergeable if they preserve the following properties:
\begin{itemize}
  \item Elements present in both the merging lists will be present in
  the final list. No element is deleted unless it is deleted in at
  least one list.
  \item An element deleted in one list will not reappear in the merged
  list (unlike Amazon's replicated shopping cart~\cite{dynamo}).
  \item The order of elements is retained. That is, if $x$ occurs
  before $y$ in one of the lists, and if both elements are present in
  the merged list, then $x$ occurs before $y$ in the merged list.
\item If an element $y$ is substituted for an element $x$ in one list,
  then $y$ is also substituted for $x$ in the merged list (i.e., $x$
  is absent and $y$ is present), unless $x$ is deleted or substituted
  in the other list. In the former case, deletion wins, and neither
  $x$ nor $y$ occur in the merged list. In the latter case, if $x$ is
  substituted by a different element $z$ in the other list, then the
  merged list substitutes $x$ by a merge of $y$ and $z$, defined as
  per the merge semantics of the list element type.
\end{itemize}
Thus, lists serve as an example of a polymorphic mergeable data type that
demonstrates how mergeability can be composed.  Compositionality is
reified in the the signature of a mergeable list library as shown in
Fig.~\ref{fig:mlist-sig}; observe that the type of items in the list
(\C{A.t}) is also required to be mergeable (\C{A.merge}).

The merge operation for lists is composed of two separate functions,
\C{edit\_seq} and \C{op\_transform}. Both these functions have been
implemented in less than 50 lines of OCaml included in the appendix.
The functions are described below.
\begin{itemize}
	\item \C{edit\_seq} takes a pair of lists, \C{v} and $\C{v'}$, and computes
		the shortest sequence of list operations that need to be applied on \C{v}
		to obtain \C{v'}. Such a sequence is called an \emph{edit sequence}. The
		length of the sequence corresponds to the standard notion of the \emph{edit
		distance} between the two lists, which can be computed in polynomial
		time~\cite{wagner-fischer}.  \C{edit\_seq} modifies one such algorithm to
		also compute the edit sequence (an \C{edit list}). The implementation
		specified in Fig.~\ref{fig:mlist-sig} represents edits using the type
		\C{edit}.  Constructors \C{I}, \C{D}, and \C{S} stand for \C{insert},
		\C{delete}, and \C{subst}, respectively.\footnote{ Note that \C{I}, \C{D}
		and \C{S} are \emph{not} effects; no list function generates them, and
		they are not exchanged between concurrent threads. They are simply tags
		used by the list merge operation for convenience.  Indeed, to better
		appreciate the virtues of mergeable types, we encourage the reader to
		think about how one might implement a replicated list with this
		functionality using an explicit effect representation as described in
		Sec.~\ref{sec:intro}.}  The subst constructor also carries the \C{A.t}
		element that was substituted. The element serves as the \C{lca} argument to
		\C{A.merge} in case of concurrent conflicting substitutions.
		Fig.~\ref{fig:list-eg} illustrates edit sequences for a sample list.  The
		sequence \C{[I(c,0); S(3,c,s)]} maps the list \C{[a;b;c]} to \C{[c;a;b;s]}
		because \C{subst 3 s (insert c 0 [a;b;c]) = [c;a;b;s]}.
	\item  \C{op\_transform} takes a pair of edit sequences, $s_1$ and $s_2$,
		that map a list $v$ to two different lists, $v_1$ and $v_2$ (e.g.,
		Fig.~\ref{fig:list-eg}), and transforms $s_1$ to $s_1'$ such that $s_1'$
		has the same effect on $v_2$ as $s_1$ had on $v$.  For instance, in
		Fig.~\ref{list-eg}, let $s_1$ be the edit sequence \C{[D(1); S(1,c,d)]},
		which maps the list \C{[a;b;c]} ($v$) to \C{[a;d]} ($v_1$). The \C{D} edit
		deletes the first element of $v$. However, if $s_1$ is to be applied on the
		list \C{[c;a;b;s]} ($v_2$), which has a new element at position 0, the
		\C{D} edit must delete the the second element from $v_2$, if it were to
		have the same effect as it did on $v$. We say \C{D(1)} in $s_1$ has to be
		transformed w.r.t the sequence $s_2 = \C{[I(c,0);S(3,c,s)]}$.  The function
		\C{op\_transform} (Fig.~\ref{fig:mlist-sig}) computes such
		\emph{operational transformations} for an edit sequence $s_1$ w.r.t another
		edit sequence $s_2$. A notable transformation rule is for a substitution
		w.r.t another substitution at the same position. Since the substituting
		items could be different, the function relies on the item merge function
		(\C{A.merge}) to merge them into a single item. For instance, in
		Fig.~\ref{fig:list-eg}, there are simultaneous substitutions to the item
		\C{c} in $s_1$ and $s_2$; while $s_1$ substitutes it with \C{d}, $s_2$ does
		it with \C{s}. The merged list therefore substitutes \C{c} with the merge
		of \C{d} and \C{s}, defined as per the merge semantics of \C{A.t}.
\end{itemize}
The \C{MList.merge} can now be defined as a function that first
computes edit sequences, $s_1$ and $s_2$, for the two concurrent
lists, $v_1$ and $v_2$ w.r.t their common ancestor $v$, transforms one
of them (say, $s_1$) w.r.t $s_2$ to obtain $s_1'$, and finally applies
(the operations defined by) $s_1'$ to $v_2$ to obtain the merged list.

>>>>

\input{rope-sig}

@@

Fig.~\ref{fig:mrope-sig} shows a rope interface in OCaml. The type of
the list of values is left abstract (\C{A.t}); it can be instantiated
with any module that supports \C{concat} and \C{split} operations. The
interface shown in Fig.~\ref{fig:mrope-sig} is that of a mergeable
rope (\C{MRope}), whose \C{merge} implementation is shown
in Fig.~\ref{fig:mrope-merge}. The function returns one of the
versions (\C{l} or \C{r}), if the other one is same as the common
ancestor. Otherwise, it calls \C{merge\_rec}, which recursively
compares the structures of merging ropes, and their common ancestor.
The function recurses until one of the ropes differs from the other
two, or all of them are leaf nodes, at which point it relies on the
(abstract) list merge function (\C{A.merge}) to merge the lists
represented by the ropes. \C{merge\_rec} relies on auxiliary
functions, \C{mk\_rope} and \C{create\_index}, whose definitions are
not shown. The former constructs a rope given a list (\C{A.t}), and
the latter computes the index at a node, given its left sub-tree.

One feature of the mergeable rope implementation in
Fig.~\ref{fig:rope-sig}, that is characterisic of mergeable types in
general, is compositionality; rope is a polymorphic mergeable data
type that composes mergeability. The compositionality is concretized
by the \C{MRope} signature , which requires the type of items in the
list (\C{A.t}) to also be mergeable (\C{A.merge}).

@@

{\bf Lists}. As an abstract data type, lists support three
operations: \C{insert x i l}, to insert an element \C{x} at position
\C{i} in the list \C{l}, \C{delete i l}, to delete the element at
\C{i}'th position, and \C{subst i x l}, to substitute the element at
the \C{i}'th position with \C{x}.  Lists of mergeable items are
mergeable if they preserve the following properties:
\begin{itemize}
  \item Elements present in both the merging lists will be present in
  the final list. No element is deleted unless it is deleted in at
  least one list.
  \item An element deleted in one list will not reappear in the merged
  list (unlike Amazon's replicated shopping cart~\cite{dynamo}).
  \item The order of elements is retained. That is, if $x$ occurs
  before $y$ in one of the lists, and if both elements are present in
  the merged list, then $x$ occurs before $y$ in the merged list.
\item If an element $y$ is substituted for an element $x$ in one list,
  then $y$ is also substituted for $x$ in the merged list (i.e., $x$
  is absent and $y$ is present), unless $x$ is deleted or substituted
  in the other list. In the former case, deletion wins, and neither
  $x$ nor $y$ occur in the merged list. In the latter case, if $x$ is
  substituted by a different element $z$ in the other list, then the
  merged list substitutes $x$ by a merge of $y$ and $z$, defined as
  per the merge semantics of the list element type.
\end{itemize}

The merge operation for lists is composed of two separate functions,
\C{edit\_seq} and \C{op\_transform}. Both these functions have been
implemented in less than 50 lines of OCaml included in the appendix.
The functions are described below.
\begin{itemize}
	\item \C{edit\_seq} takes a pair of lists, \C{v} and $\C{v'}$, and computes
		the shortest sequence of list operations that need to be applied on \C{v}
		to obtain \C{v'}. Such a sequence is called an \emph{edit sequence}. The
		length of the sequence corresponds to the standard notion of the \emph{edit
		distance} between the two lists, which can be computed in polynomial
		time~\cite{wagner-fischer}.  \C{edit\_seq} modifies one such algorithm to
		also compute the edit sequence (an \C{edit list}). The implementation
		specified in Fig.~\ref{fig:mlist-sig} represents edits using the type
		\C{edit}.  Constructors \C{I}, \C{D}, and \C{S} stand for \C{insert},
		\C{delete}, and \C{subst}, respectively.\footnote{ Note that \C{I}, \C{D}
		and \C{S} are \emph{not} effects; no list function generates them, and
		they are not exchanged between concurrent threads. They are simply tags
		used by the list merge operation for convenience.  Indeed, to better
		appreciate the virtues of mergeable types, we encourage the reader to
		think about how one might implement a replicated list with this
		functionality using an explicit effect representation as described in
		Sec.~\ref{sec:intro}.}  The subst constructor also carries the \C{A.t}
		element that was substituted. The element serves as the \C{lca} argument to
		\C{A.merge} in case of concurrent conflicting substitutions.
		Fig.~\ref{fig:list-eg} illustrates edit sequences for a sample list.  The
		sequence \C{[I(c,0); S(3,c,s)]} maps the list \C{[a;b;c]} to \C{[c;a;b;s]}
		because \C{subst 3 s (insert c 0 [a;b;c]) = [c;a;b;s]}.
	\item  \C{op\_transform} takes a pair of edit sequences, $s_1$ and $s_2$,
		that map a list $v$ to two different lists, $v_1$ and $v_2$ (e.g.,
		Fig.~\ref{fig:list-eg}), and transforms $s_1$ to $s_1'$ such that $s_1'$
		has the same effect on $v_2$ as $s_1$ had on $v$.  For instance, in
		Fig.~\ref{list-eg}, let $s_1$ be the edit sequence \C{[D(1); S(1,c,d)]},
		which maps the list \C{[a;b;c]} ($v$) to \C{[a;d]} ($v_1$). The \C{D} edit
		deletes the first element of $v$. However, if $s_1$ is to be applied on the
		list \C{[c;a;b;s]} ($v_2$), which has a new element at position 0, the
		\C{D} edit must delete the second element from $v_2$, if it were to
		have the same effect as it did on $v$. We say \C{D(1)} in $s_1$ has to be
		transformed w.r.t the sequence $s_2 = \C{[I(c,0);S(3,c,s)]}$.  The function
		\C{op\_transform} (Fig.~\ref{fig:mlist-sig}) computes such
		\emph{operational transformations} for an edit sequence $s_1$ w.r.t another
		edit sequence $s_2$. A notable transformation rule is for a substitution
		w.r.t another substitution at the same position. Since the substituting
		items could be different, the function relies on the item merge function
		(\C{A.merge}) to merge them into a single item. For instance, in
		Fig.~\ref{fig:list-eg}, there are simultaneous substitutions to the item
		\C{c} in $s_1$ and $s_2$; while $s_1$ substitutes it with \C{d}, $s_2$ does
		it with \C{s}. The merged list therefore substitutes \C{c} with the merge
		of \C{d} and \C{s}, defined as per the merge semantics of \C{A.t}.
\end{itemize}
The \C{MList.merge} can now be defined as a function that first
computes edit sequences, $s_1$ and $s_2$, for the two concurrent
lists, $v_1$ and $v_2$ w.r.t their common ancestor $v$, transforms one
of them (say, $s_1$) w.r.t $s_2$ to obtain $s_1'$, and finally applies
(the operations defined by) $s_1'$ to $v_2$ to obtain the merged list.
